### 2018-06-11 express2018/ccs.tex
Cumulative updates (still not touching the 2nd part)

\finish{maybe use recursion in place of constants, so to be
  more faithful wrt the HOL formalisation
}

We assume  an infinite set  of \emph{names} $  a, b, \ldots $
 and   a  set  of
\emph{constant identifiers} (or simply \emph{constants}) 
 for writing recursive processes. 
The special symbol $\tau$ does not occur in the names and in the 
constants. 
The class   of the  CCS processes  is
built from the operators
of  parallel
composition,  sum,
 restriction, input and output prefix ???,  
and  constants:
$$ \begin{array}{ccl}
P  & := &    \outC a . P \midd \; a . P \midd \;  P_1 |  P_2 \; \midd  \;
P_1 + P_2 \; \midd % \; \mu . P\; \midd  \; 
  \res a P  \;  \midd \;   K  \;  \midd \; \nil  \\[\tkpS]
   \end{array}
 $$
Each constant $K$ has  a definition 
 $K \Defi P$.
We sometimes omit trailing $\nil$, e.g., writing $a|b$ for $a.\nil |b .\nil $ .
The operational semantics is given by means of an LTS, and is
reported in Figure~\ref{f:LTSCCS} (the symmetric version  of the two rules for
parallel composition and the rule for sum  has been omitted).

 The \emph{immediate derivatives} of a
process $P$ are the elements of  the set $\{P' \st P \arr\mu P' \mbox{ for some $\mu$ }
\}$.   
We use $\ell$ to range over
 visible actions (i.e.,  inputs or outputs, excluding  $\tau$).
\begin{figure*}
\begin{center}
 $\displaystyle{   P \arr\mu   P' \over   P + Q   \arr\mu
P'  } $  $ \hb$   
%{{\sf par1}}:
 $\displaystyle{   P \arr\mu   P' \over   P | Q   \arr\mu
P' | Q } $  $ \hb$   
  $\; \;$  $\displaystyle{ P \arr{ a}P' \hk \hk  Q
\arr{\outC a }Q'  \over     P|  Q \arr{ \tau} P'
|  Q'  }$ 
\\
\hskip .6cm  
 $\displaystyle{  \over  \mu.  P    \arr\mu
P } $  $ \hb$   
\hskip .6cm  
$\displaystyle{ P \arr{\mu}P' \over
 \res a     P   \arr{\mu} \res a P'} $ $ \mu \neq a, \outC a$
$ \hb$ %  &
\hskip .5cm 
$\displaystyle{ P \arr{\mu}P' \over
 K   \arr{ \mu} P'  } $  if  $  K \Defi P$
\end{center}
%  \begin{center}
% \begin{tabular}{rlcrlcrl}
% %{{\sf sum}}:
% & $  \Sigma_{i\in I} \mu_i.   P_i 
% \arr{ \mu_i  }P_i$ & $ \hb$ &
% %{{\sf par1}}:
% & $\displaystyle{   P \arr\mu   P' \over   P | Q   \arr\mu
% P' | Q } $ & $ \hb$ &  
% % \multicolumn{5}{c}{
% %  {{\sf com1}}: 
% &  $\; \;$  $\displaystyle{ P \arr{ a}P' \hk \hk  Q
% \arr{\outC a }Q'  \over     P|  Q \arr{ \tau} P'
% |  Q'  }$ 
%  \\[\mysp]
% \multicolumn{8}{c}{ 
% %{{\sf res}}:  %&
% \hskip .4cm  
%  $\displaystyle{ P \arr{\mu}P' \over
%  \res a     P   \arr{\mu} \res a P'} $ $ \mu \neq a, \outC a$
% % & 
% $ \hb$ %  &
% \hskip .4cm 
% %{{\sf const}}: %& 
% \hskip .3cm 
% $\displaystyle{ P \arr{\mu}P' \over
%  K   \arr{ \mu} P'  } $  if  $  K \Defi P$
% }\end{tabular}
%  \end{center} 
\caption{The LTS for CCS}
\label{f:LTSCCS}
\end{figure*}

>>>>

\finish{maybe use recursion in place of constants, so to be
  more faithful wrt the HOL formalization}

We assume  a possibly infinite set of \emph{input names} $a, b,
\ldots$ and \emph{output names} $\outC a, \outC b, \ldots$,
 and a set of \emph{constant identifiers} (or simply \emph{constants})
 $A, B, \ldots$ for writing recursive processes. 
The special symbol $\tau$ does not occur in the names and in the 
constants. 
The class  of the CCS processes is inductively built from $\nil$ by the operators
of parallel composition,  sum,
 restriction, prefix, and constants:
\begin{equation*}
\begin{array}{ccl}
P  & := &  \nil \; \midd \;  \mu . P \; \midd \;  P_1 |  P_2 \; \midd  \;
P_1 + P_2 \; \midd % \; \mu . P\; \midd  \; 
  (\res a\!)\, P  \;  \midd \;   K   %\\[\tkpS]
\end{array}
\end{equation*}
Each constant $K$ has  a definition 
 $K \Defi P$.
We sometimes omit trailing $\nil$, e.g., writing $a|b$ for $a.\nil |b .\nil $ .
The operational semantics is given by means of an LTS, and is
reported in Figure~\ref{f:LTSCCS} (the symmetric version  of the two rules for
parallel composition and the rule for sum  has been omitted).

 The \emph{immediate derivatives} of a
process $P$ are the elements of  the set $\{P' \st P \arr\mu P' \mbox{ for some $\mu$ }
\}$.   
We use $\ell$ to range over
 visible actions (i.e.,  inputs or outputs, excluding  $\tau$).
\begin{figure*}
\begin{center}
 $\displaystyle{   P \arr\mu   P' \over   P + Q   \arr\mu
P'  } $  $ \hb$   
%{{\sf par1}}:
 $\displaystyle{   P \arr\mu   P' \over   P | Q   \arr\mu
P' | Q } $  $ \hb$   
  $\; \;$  $\displaystyle{ P \arr{ a}P' \hk \hk  Q
\arr{\outC a }Q'  \over     P|  Q \arr{ \tau} P'
|  Q'  }$ 
\\
\vskip .1cm
\hskip .6cm  
 $\displaystyle{  \over  \mu.  P    \arr\mu
P } $  $ \hb$   
\hskip .6cm  
$\displaystyle{ P \arr{\mu}P' \over
 (\res a\!)\, P   \arr{\mu} (\res a\!)\, P'} $ $ \mu \neq a, \outC a$
$ \hb$ %  &
\hskip .5cm 
$\displaystyle{ P \arr{\mu}P' \over
 K   \arr{ \mu} P'  } $  if  $  K \Defi P$
\end{center}
%  \begin{center}
% \begin{tabular}{rlcrlcrl}
% %{{\sf sum}}:
% & $  \Sigma_{i\in I} \mu_i.   P_i 
% \arr{ \mu_i  }P_i$ & $ \hb$ &
% %{{\sf par1}}:
% & $\displaystyle{   P \arr\mu   P' \over   P | Q   \arr\mu
% P' | Q } $ & $ \hb$ &  
% % \multicolumn{5}{c}{
% %  {{\sf com1}}: 
% &  $\; \;$  $\displaystyle{ P \arr{ a}P' \hk \hk  Q
% \arr{\outC a }Q'  \over     P|  Q \arr{ \tau} P'
% |  Q'  }$ 
%  \\[\mysp]
% \multicolumn{8}{c}{ 
% %{{\sf res}}:  %&
% \hskip .4cm  
%  $\displaystyle{ P \arr{\mu}P' \over
%  \res a     P   \arr{\mu} \res a P'} $ $ \mu \neq a, \outC a$
% % & 
% $ \hb$ %  &
% \hskip .4cm 
% %{{\sf const}}: %& 
% \hskip .3cm 
% $\displaystyle{ P \arr{\mu}P' \over
%  K   \arr{ \mu} P'  } $  if  $  K \Defi P$
% }\end{tabular}
%  \end{center} 
\caption{The LTS of CCS}
\label{f:LTSCCS}
\end{figure*}

### 2018-06-11 express2018/expansion.tex
Cumulative updates (still not touching the 2nd part)

\item the converse of (1)
 on the actions from $Q$.
%\item if $Q  \Ar\mu   Q'$, then there is $P'$ s.t.\ $P
%\Arcap\mu  P'$ and $P' 
%\RR Q'$.
\end{enumerate}  
 $P$ and $Q$ are \emph{bisimilar},
written $P \wb
 Q$, if $P \RR Q$ for some  bisimulation $\R$.  
\qed\end{definition} 
\finish{the definition of rooted bisimilarity }

We sometimes call bisimilarity \emph{weak} bisimilarity, to
distinguish it from \emph{strong bisimilarity}, $\sim$,
obtained by replacing in the above definition   the weak answer $
Q\Arcap\mu Q'$ with the strong  $Q \arr \mu Q'$.

>>>>

\item the converse of (1)
 on the actions from $Q$.
%\item if $Q  \Ar\mu   Q'$, then there is $P'$ s.t.\ $P
%\Arcap\mu  P'$ and $P' 
%\RR Q'$.
\end{enumerate}  
 $P$ and $Q$ are \emph{bisimilar},
written $P \wb
 Q$, if $P \RR Q$ for some  bisimulation $\R$.  
\end{definition}

\begin{definition}{rooted bisimilarity}

\end{definition}
\finish{the definition of rooted bisimilarity }

We sometimes call bisimilarity as ``\emph{weak} bisimilarity'', to
distinguish it from \emph{strong bisimilarity}, $\sim$,
obtained by replacing in the above definition   the weak answer $
Q\Arcap\mu Q'$ with the strong  $Q \arr \mu Q'$.

### 2018-06-11 express2018/intro.tex
Added (candidate) title and abstract
Cumulative updates (still not touching the 2nd part)

A prominent proof method for bisimulation, put forward by Milner and widely used in his
 landmark  CCS book \cite{Mil89}
is
\emph{unique solution of equations}, whereby two tuples of processes are
componentwise bisimilar if they are solutions 
of the same system of equations.
This method  is important in verification techniques and tools
  based on 
algebraic      reasoning \cite{theoryAndPractice,RosUnder10,BaeBOOK}.

In the   \emph{weak} case (when  behavioural equivalences abstract from internal moves,
which practically is the most relevant case), however, 
Milner's proof method has severe syntactical limitations. 
To overcome such limitations, Sangiorgi proposes to replace
equations with  special inequations called
\emph{contractions} \cite{...}. Contraction is a preorder that, roughly, places some efficiency
constraints on processes.  
Uniqueness of the solutions 
 of a system of contractions
 is defined as with systems of
equations:  
any two solutions must be bisimilar.
The difference with equations is in the meaning of solution:
in the case of contractions
the solution is evaluated with respect to
the contraction preorder, rather than bisimilarity. 
With contractions, most syntactic limitations of the unique-solution theorem can be
removed.  One constraint that still remains in \cite{San...} is on occurrences of the sum operator, due to
the failure of substitutivity of contraction wrt such operator.

The purpose of this paper is twofold. 
On the one hand, we refine the contraction proof technique by moving to the 
\emph{rooted contraction}, that is, the largest congruence included in the contraction
preoder.  Using rooted contraction one obtains a unique-solution theorem that is valid for
\emph{rooted bisimilarity} (hence also for bisimilarity itself).  
On the other hand, we provide a  
 comprehensive formalization  of the core of the theory of CCS 
 in the HOL
theorem prover (HOL4). The formalisation  includes the theory of unique solution of equations and
contractions discussed in the present paper, but is not limited to it (partly because such
theorems rely on a number of more fundamental results):
indeed the formalisation encompasses the basic properties of strong and weak
bisimilarity (e.g. the fixed-point and substitutivity properties), 
their algebraic theory, 
 various
versions of ``bisimulation up to'' techniques (e.g., bisimulation up-to expansion),
the basic properties  of rooted bisimilarity. Considering the relationship between bisimilarity and rooted
bisimilarity, the formalisation includes the proof that the latter is the coarsest
congruence included in the former, for which two proofs are formalised: one as in Milner's
book,  requiring the hypothesis that  no processes can use all labels; the other without
such hypothesis, essentially formalising van Glabbek's paper ...
Similar theorems are proved for rooted contractions wrt the contraction preorder.

\finish{for Express, insist more on structured operational semantics,
  to fit the scope of the conference}

>>>>

A prominent proof method for bisimulation, put forward by Milner and widely used in his
landmark CCS book \cite{Mil89} is the
\emph{unique solution of equations}, whereby two tuples of processes are
componentwise bisimilar if they are solutions 
of the same system of equations.
This method  is important in verification techniques and tools
based on algebraic reasoning \cite{theoryAndPractice,RosUnder10,BaeBOOK}.

In the   \emph{weak} case (when  behavioural equivalences abstract from internal moves,
which practically is the most relevant case), however, 
Milner's proof method has severe syntactical limitations. 
To overcome such limitations, Sangiorgi proposes to replace
equations with  special inequations called
\emph{contractions} \cite{sangiorgi2015equations}. Contraction is a
preorder that, roughly, places some efficiency
constraints on processes.  Uniqueness of the solutions of a system of contractions
 is defined as with systems of equations:  
any two solutions must be bisimilar.
The difference with equations is in the meaning of solution:
in the case of contractions
the solution is evaluated with respect to
the contraction preorder, rather than bisimilarity. 
With contractions, most syntactic limitations of the unique-solution theorem can be
removed.  One constraint that still remains in
\cite{sangiorgi2015equations} is on occurrences of the sum operator,
due to the failure of substitutivity of contraction w.r.t. such operator.

The purpose of this paper is twofold. 
On the one hand, we refine the contraction proof technique by moving to the 
\emph{rooted contraction} \cite{Tian:2017wrba}, that is, the coarsest (pre)congruence contained in the contraction
preorder.  Using rooted contraction one obtains an unique-solution theorem that is valid for
\emph{rooted bisimilarity} (hence also for bisimilarity itself).  
On the other hand, we provide a  
 comprehensive formalization  of the core of the theory of CCS 
 in the HOL
theorem prover (HOL4). The formalization  includes the proofs of
Milner's 3 ``unique solution of equations'' theorems and
contractions discussed in the present paper, but is not limited to it (partly because such
theorems rely on a number of more fundamental results):
indeed the formalization encompasses the basic properties of strong and weak
bisimilarity (e.g. the fixed-point and substitutivity properties), 
their algebraic theory, various versions of ``bisimulation up to''
techniques (e.g., bisimulation up-to expansion),
the basic properties  of rooted bisimilarity. Considering the relationship between bisimilarity and rooted
bisimilarity, the formalization includes the proof that the latter is the coarsest
congruence included in the former, for which two proofs are formalized: one as in Milner's
book,  requiring the hypothesis that  no processes can use all labels; the other without
such hypothesis, essentially formalising van Glabbek's paper \cite{vanGlabbeek:2005ur}.
Similar theorems are proved for rooted contractions wrt the contraction preorder.

\finish{for Express, insist more on Structural Operational Semantics,
  to fit the scope of the conference}
